---
title: 'Reproducible Research: Peer Assessment 1'
author: "Roberto Rivera"
date: "04/26/2015"
output:
  html_document:
    keep_md: yes
  pdf_document: default
geometry: margin=0.75in
---
\fontsize{9}{10}

## Loading and preprocessing the data

This section involves:  

 - downloading, unzipping and mapping the data  
 
 - creating a date-time field and   
 
 - creating a couple of aggregated data sets.  

```{r echo=TRUE, eval=TRUE, cache=TRUE}
# The variables included in this dataset are:
# * steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
# * date: The date on which the measurement was taken in YYYY-MM-DD format
# * interval: Identifier for the 5-minute interval in which measurement was taken
# The dataset is stored in a comma-separated-value (CSV) file and there are a total of 
# 17,568 observations in this dataset.
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipFile = "./activity.zip"
download.file(fileUrl, destfile = zipFile, method = "curl")

# check file
dateDownloaded <- date(); dateDownloaded
unzip(zipFile, exdir="./")
```

```{r echo=TRUE, eval=TRUE, cache=TRUE}
library(reshape2)
library(dplyr)
library(pander)
library(stringr)

# read activity data
dat.a <- read.csv("activity.csv")
# add date time fields
dat.a$dttime <- str_pad(dat.a$interval, 4, pad = "0")
dat.a$dttime <- as.POSIXct(str_c(dat.a$date, " ", dat.a$dttime), format = "%Y-%m-%d %H%M")

# create date summary
dat.summ <- dat.a[,c(1:2)] %>%
    filter(steps > 0) %>%
    arrange(date, steps) %>%
    group_by(date) %>% 
    summarise(Total=as.numeric(sum(steps)), 
              Mean=round(mean(steps),0), 
              Median=round(median(steps),0)) %>%
    data.frame()
dat.summ$date <- as.Date(dat.summ$date)

# create date average summary
dat.avg <- dat.summ %>%
    melt(id="date") %>%
    filter(variable %in% c("Mean","Median"))

# create interval summary
dat.intv <- dat.a[,c(1,3)] %>%
    filter(steps > 0) %>%
    arrange(interval, steps) %>%
    group_by(interval) %>% 
    summarise(Total=as.numeric(sum(steps)), 
              Mean=round(mean(steps),0), 
              Median=round(median(steps),0)) %>%
    data.frame()

```

## What is mean total number of steps taken per day?

My calculations account for intervals having more than zero steps. See the results in the plots below. 

```{r echo=TRUE, eval=TRUE, cache=TRUE}
library(ggplot2)
library(gridExtra)

grid.arrange(
    ggplot(dat.summ, aes(date)) + #facet_grid(. ~ variable) + 
        geom_histogram(stat="identity", alpha=0.2, aes(y=Total)) +
        ylab("Steps") + xlab("Date") + theme_bw() + 
        theme(axis.text.x=element_text(angle=90)) +
        scale_x_date(breaks = "1 week") +
        ggtitle("Total Steps Taken Per Day"),
    ggplot(dat.avg, aes(x=date, y=value, color=variable)) + #facet_grid(. ~ variable) + 
        geom_histogram(stat="identity", alpha=0.2, aes(y=value)) +
        ylab("Steps") + xlab("Date") + theme_bw() + 
        theme(axis.text.x=element_text(angle=90)) +
        scale_x_date(breaks = "1 week") +
        ggtitle("Average Steps Taken Per Day"),
    ncol=2)
```

## What is the average daily activity pattern?

On average the maximum number of steps occur around 8:30am.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
plot(dat.intv$interval, dat.intv$Mean, type="l", xaxt="n",
     xlab= "Five Minute Interval", ylab= "Average Steps", col="orange" , lwd=2)
axis(1, xaxp=c(0, 2400, 24), las=3)
grid(nx=24)
```

## Imputing missing values

***There are 2,304 missing step values across the data set.***

```{r echo=TRUE, eval=TRUE, cache=TRUE}
library(VIM);
panderOptions('table.split.table', Inf)
pander(md.pattern(dat.a))         # identify missig data pattern
```

***To impute the missing step values, I used the MICE (Multivariate Imputation by Chained Equations) package. This method seemed to be rather straightforward and clear.***

```{r echo=TRUE, eval=TRUE, cache=TRUE}
library(mice)
dat.imp <- mice(dat.a)            # generate imputed data sets
dat.imp <- complete(dat.imp)      # combine imputed data set
dat.imp$date   <- as.Date(dat.imp$date)
```

***These plots indicate how the step values look before and after imputation:***

```{r echo=TRUE, eval=TRUE, cache=TRUE}
par(mfrow = c(1, 2));
pbox(dat.a, pos = 2)              # plot missing data
pbox(dat.imp, pos = 2)
par(mfrow = c(1, 1));
```

## Are there differences in activity patterns between weekdays and weekends?

When comparing the number of steps taken by weekend and weekday, there are a few interesting observations. The total number of steps appear to fluctuate more during weekdays.  In addition, the time of day when steps are taken vary by weekday or weekend.  You can see that more walking happens around 1am on the weekends for example.

```{r echo=TRUE, eval=TRUE, cache=TRUE}
dat.imp$wk_day <- weekdays(dat.imp$date)
dat.imp$wkd_in <- as.factor(
    ifelse(dat.imp$wk_day %in% c("Saturday","Sunday"), "Wk_End", "Wk_Day"))

# create interval summary
dat.int2 <- dat.imp[,c(1,3,6)] %>%
    filter(steps > 0) %>%
    arrange(interval, steps) %>%
    group_by(interval, wkd_in) %>% 
    summarise(Total=as.numeric(sum(steps)), 
              Mean=round(mean(steps),0), 
              Median=round(median(steps),0)) %>%
    data.frame()

ggplot(dat.imp, aes(date, color=wkd_in)) + #facet_grid(. ~ wkd_in) + 
    geom_histogram(stat="identity", alpha=0.2, aes(y=steps)) +
    ylab("Steps") + xlab("Date") + theme_bw() + 
    theme(axis.text.x=element_text(angle=90)) +
    ggtitle("Total Steps Taken Per Day")
ggplot(data = dat.int2, aes(x=interval, y=Mean)) + geom_line(aes(colour=wkd_in)) +
    theme(axis.text.x=element_text(angle=90)) +
    scale_x_continuous(breaks = c(seq(from = 0, to = 2400, by = 100)))

```


